---
title: "Risk Taxonomy"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk Taxonomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
library(BORG)
```

This document catalogs all evaluation risks that BORG detects, organized by
severity and mechanism. Each risk type includes a description of what triggers
it, why it matters for real-world modelling, and how to fix it.

## Risk Classification

BORG classifies risks into two categories based on their impact on evaluation
validity:

| Category | Impact | BORG Response |
|----------|--------|---------------|
| **Hard Violation** | Results are invalid | Blocks evaluation, requires fix |
| **Soft Inflation** | Results are biased | Warns, allows with caution |

Hard violations indicate that the evaluation result is not trustworthy at all.
The computed metrics could be arbitrarily optimistic, and no conclusion should
be drawn from them. Soft inflation risks mean the metrics are likely
overestimated, but the direction of model comparisons may still hold. Fixing
soft risks is best practice; fixing hard violations is mandatory.


# Hard Violations

These make your evaluation results invalid. Any metrics computed with these
violations are unreliable.

## 1. Index Overlap

**What**: Same row indices appear in both training and test sets.

**Why it matters**: Because the model has seen the exact data it is being
tested on, any sufficiently flexible learner will achieve perfect scores on the
overlapping portion of the test set. In practice, index
overlap usually arises from manual bookkeeping errors: copying indices
incorrectly, using inclusive rather than exclusive ranges, or inadvertently
sharing a row between two data splits. The bug is easy to introduce and hard
to catch by looking at metrics alone, because a small overlap (say 5% of the
test set) produces a subtle, plausible-looking inflation.

**Detection**: Set intersection of `train_idx` and `test_idx`.

```{r index-overlap}
data <- data.frame(x = 1:100, y = rnorm(100))

# Accidental overlap: indices 51-60 appear in both sets
result <- borg_inspect(data, train_idx = 1:60, test_idx = 51:100)
result
```

**Fix**: Ensure indices are mutually exclusive. Use `setdiff()` to remove any
shared indices, or generate non-overlapping splits from the start with
`sample()` and complementary indexing.

## 2. Duplicate Rows

**What**: Test set contains rows identical to training rows.

**Why it matters**: Disjoint indices do not guarantee distinct data. The
underlying rows may still be exact copies. This happens often in observational
datasets that were merged from multiple sources (where the same record appears
in two files), in datasets with many categorical predictors and few unique
combinations, or when data augmentation accidentally copies real rows into a
held-out set. A model that memorizes training patterns will score well on these
duplicates regardless of its true generalization ability. The effect is
proportional to the fraction of test rows that are duplicates: 10% duplicates
in the test set can inflate accuracy by several percentage points.

**Detection**: Row hashing and comparison (C++ backend for numeric data).

```{r duplicate-rows}
# Data with duplicate rows
dup_data <- rbind(
  data.frame(x = 1:5, y = 1:5),
  data.frame(x = 1:5, y = 1:5)  # Duplicates
)

result <- borg_inspect(dup_data, train_idx = 1:5, test_idx = 6:10)
result
```

**Fix**: Remove duplicate rows before splitting, or ensure splits respect
duplicates by keeping all copies in the same set. In ecological survey data,
duplicates sometimes represent legitimate repeated measurements; in that case,
treat them as a group and use group-aware splitting.

## 3. Preprocessing Leakage

**What**: Normalization, imputation, or dimensionality reduction fitted on the
full dataset before splitting.

**Why it matters**: Column means, standard deviations, or PCA loadings
computed from the entire dataset carry test-set information into the training
pipeline. Information flows backwards from test to train through these shared statistics. The effect
is typically small for large datasets with independent observations (a few
percent inflation in R-squared) but can be severe for small datasets, datasets
with strong distributional shifts between train and test, or high-dimensional
data where PCA loadings are heavily sample-dependent. This is one of the most
common forms of leakage in published studies because many preprocessing
pipelines default to fitting on all available data.

**Detection**: Recompute statistics on train-only data and compare to stored
parameters. Discrepancy indicates leakage.

**Supported objects**:

| Object Type | Parameters Checked |
|-------------|-------------------|
| `caret::preProcess` | `$mean`, `$std` |
| `recipes::recipe` | Step parameters after `prep()` |
| `prcomp` | `$center`, `$scale`, rotation matrix |
| `scale()` attributes | `center`, `scale` |

```{r preprocessing-leak, eval=FALSE}
# BAD: Scale fitted on all data
scaled_data <- scale(data)  # Uses all rows!
train <- scaled_data[1:70, ]
test <- scaled_data[71:100, ]

# BORG detects this
borg_inspect(scaled_data, train_idx = 1:70, test_idx = 71:100)
```

**Fix**: Fit preprocessing on training data only, then apply to test:

```r
train_data <- data[1:70, ]
test_data <- data[71:100, ]

# Fit on train
means <- colMeans(train_data)
sds <- apply(train_data, 2, sd)

# Apply to both
train_scaled <- scale(train_data, center = means, scale = sds)
test_scaled <- scale(test_data, center = means, scale = sds)
```

## 4. Target Leakage (Direct)

**What**: Feature has absolute correlation > 0.99 with target.

**Why it matters**: Near-perfect correlation (|r| > 0.99) between a feature
and the outcome almost always means the feature was derived from the outcome, either directly or through a
chain of deterministic transformations. Classic examples include
`days_since_diagnosis` when predicting `has_disease`, `total_spent` when
predicting `is_customer`, or aggregated future values leaked into current-period
features. These features will dominate any model, producing impressive metrics
that collapse entirely in deployment where the post-hoc information is
unavailable. Target leakage is particularly common in tabular datasets assembled
by joining multiple tables, where a downstream aggregate accidentally ends up
as a predictor.

**Detection**: Compute Pearson correlation of each numeric feature with target
on training data.

```{r target-leakage}
# Use borg_simulate for controlled target leakage
sim <- borg_simulate(n = 100, type = "target_leak", leak_strength = 0.99, seed = 1)

result <- borg_inspect(sim$data, train_idx = 1:70, test_idx = 71:100,
                       target = "y")
result
```

**Fix**: Remove or investigate the leaky feature. If it is a legitimate
predictor, document why correlation > 0.99 is expected and verify the feature
is available at prediction time.

## 5. Group Leakage

**What**: Same group (patient, site, species) appears in both train and test.

**Why it matters**: Latent group-level characteristics let the model learn
patient- or site-specific intercepts it cannot use on new groups. If the same patient appears in both train and test,
the model effectively learns patient-specific intercepts during training and is
rewarded for them at test time. This inflates performance because, in
deployment, the model will encounter new patients it has never seen. The
magnitude of inflation depends on the intra-class correlation (ICC): when ICC
is high (observations within a group are very similar), group leakage can
inflate R-squared by 20-50% or more. This risk is pervasive in medical data,
ecological monitoring (repeated visits to the same site), and any longitudinal
study design.

**Detection**: `borg_diagnose()` detects clustered structure and recommends
group-aware CV. At the split level, check whether any group's members appear
in both the training and test sets.

```{r group-leakage, eval=FALSE}
# Clinical data with patient IDs
clinical <- data.frame(
  patient_id = rep(1:10, each = 10),
  measurement = rnorm(100)
)

# Random split ignoring patients (groups will leak)
set.seed(123)
all_idx <- sample(100)
train_idx <- all_idx[1:70]
test_idx <- all_idx[71:100]

# Check which patients appear in both sets
train_patients <- unique(clinical$patient_id[train_idx])
test_patients <- unique(clinical$patient_id[test_idx])
shared <- intersect(train_patients, test_patients)
cat("Shared patients:", shared, "\n")
```

**Fix**: Use group-aware splitting so that entire groups are held out together:

```r
# Split at the patient level
train_patients <- sample(unique(clinical$patient_id), 7)
train_idx <- which(clinical$patient_id %in% train_patients)
test_idx <- which(!clinical$patient_id %in% train_patients)
```

Alternatively, use `borg_cv()` with the `groups` argument to generate
group-respecting folds automatically.

## 6. Temporal Ordering Violation

**What**: Test observations predate training observations.

**Why it matters**: Predicting the past from the future is look-ahead bias,
and in any time-dependent process the effect is severe. The model can exploit temporal trends,
seasonal patterns, or regime shifts that would not be available at prediction
time. This violation is especially damaging for autoregressive features (lagged
values, rolling averages), because those features were computed under the
assumption that earlier data is available. Random splitting of time series data
is one of the most common methodological errors in applied ML: it is easy to
do, produces excellent-looking metrics, and fails silently. Studies that
randomly split stock prices, weather records, or patient trajectories routinely
report inflated accuracy that disappears in live deployment.

**Detection**: Compare max training timestamp to min test timestamp. Any test
observation that falls before a training observation constitutes a violation.

```{r temporal-leak}
# Simulate temporal data
sim <- borg_simulate(n = 100, type = "temporal", autocorrelation = 0.7, seed = 1)

# Wrong: random split ignores time ordering
set.seed(42)
random_idx <- sample(100)
train_idx <- random_idx[1:70]
test_idx <- random_idx[71:100]

# borg_diagnose detects the temporal structure
diagnosis <- borg_diagnose(sim$data, time = "time", target = "y", verbose = FALSE)
diagnosis@recommended_cv
```

**Fix**: Use chronological splits where all test data comes after training:

```r
train_idx <- 1:70
test_idx <- 71:100
```

## 7. CV Fold Contamination

**What**: Cross-validation folds contain test indices, or folds overlap
incorrectly.

**Why it matters**: For nested cross-validation to be valid, the outer test
set must remain completely invisible to all inner folds. If the outer test indices leak into any
inner fold's training set, the tuning procedure has seen the test data. This is
a subtle and surprisingly common mistake when manually constructing fold
objects. It invalidates the entire nested CV result because the hyperparameter
selection was influenced by test data.

**Detection**: Check if any fold's training indices intersect with held-out test
set.

**Supported objects**:

- `caret::trainControl`: checks `$index` and `$indexOut`

- `rsample::vfold_cv` and other `rset` objects

- `rsample::rsplit` objects

## 8. Model Scope

**What**: Model was trained on more rows than the claimed training set.

**Why it matters**: Evidence inside the fitted object that it saw more rows
than the training indices specify means test data leaked into fitting. This can happen indirectly through hyperparameter tuning on the full
dataset, through ensemble methods that use the entire dataset for base learner
training, or simply through a bookkeeping mismatch between the reported split
and the actual training call. Unlike index overlap (which checks the split
definition), model scope checks the fitted model itself for signs that it saw
more data than it should have.

**Detection**: Compare `nrow(trainingData)` or `length(fitted.values)` to
`length(train_idx)`.

**Supported objects**: `lm`, `glm`, `ranger`, `caret::train`, parsnip models,
workflows.


# Soft Inflation Risks

These bias results but may not completely invalidate them. Model ranking might
be preserved even if absolute metrics are optimistic.

## 1. Target Leakage (Proxy)

**What**: Feature has correlation 0.95-0.99 with target.

**Why warning not error**: A correlation in this range can represent either a
legitimately strong predictor or a subtler form of target leakage. Domain
knowledge is required to distinguish the two cases. For example, in ecological
modelling, mean annual temperature might genuinely correlate at r = 0.97 with
species richness. Conversely, a feature like `average_daily_sales_last_year`
might correlate at r = 0.96 with `annual_revenue` because it is a near-direct
transformation of the target. BORG warns but does not block, giving you the
opportunity to investigate.

**Detection**: Same as direct leakage, different threshold.

```{r proxy-leakage}
# Use borg_simulate for controlled proxy leakage
sim <- borg_simulate(n = 100, type = "target_leak", leak_strength = 0.7, seed = 1)

result <- borg_inspect(sim$data, train_idx = 1:70, test_idx = 71:100,
                       target = "y")
result
```

**Action**: Review whether the feature should be available at prediction time
in production. If it is a legitimate predictor, document the justification.

## 2. Spatial Proximity

**What**: Test points are very close to training points in geographic space.

**Why it matters**: Nearby locations share environmental conditions, soil
properties, and species compositions through spatial autocorrelation. When train
and test points are spatially interleaved, the model can interpolate between
nearby training points rather than learning generalizable patterns. Error estimates become optimistic. The magnitude scales with the strength of
spatial autocorrelation and the ratio of inter-point distance to autocorrelation
range. In ecological and environmental modelling, where spatial structure is the
rule rather than the exception, this inflation can be severe: reported R-squared
values may drop by 30-60% when switching from random to spatially blocked CV.

**Detection**: Compute minimum distance from each test point to nearest
training point. Flag if < 1% of spatial spread.

```{r spatial-proximity}
# Use borg_simulate for spatially autocorrelated data
sim <- borg_simulate(n = 100, type = "spatial", autocorrelation = 0.7, seed = 42)

# Random split intermixes nearby points
set.seed(42)
train_idx <- sample(100, 70)
test_idx <- setdiff(1:100, train_idx)

result <- borg_inspect(sim$data, train_idx = train_idx, test_idx = test_idx,
                       coords = c("x", "y"))
result
```

**Fix**: Use spatial blocking so that train and test occupy distinct geographic
regions:

```r
# Geographic split along x-axis
train_idx <- which(sim$data$x < 0.5)
test_idx <- which(sim$data$x >= 0.5)
```

## 3. Spatial Overlap

**What**: Test region falls inside training region's convex hull.

**Why it matters**: Because interpolation is easier than extrapolation, test points
are surrounded by training points, the model benefits from having nearby
reference data in all directions. In deployment, the model may need to predict
in regions where no training data exists (true extrapolation). Performance on
interpolated test points overestimates the model's transferability to new
areas.

**Detection**: Compute convex hull of training points, count test points
inside.

**Threshold**: Warning if > 50% of test points fall inside training hull.

## 4. Random CV on Dependent Data

**What**: Using random k-fold CV when data has spatial, temporal, or group
structure.

**Why it matters**: Breaking dependencies artificially, random folds produce
optimistic error estimates. When observations are not independent (e.g.,
spatial neighbors, repeated measures, time series), random CV treats each
observation as if it were drawn independently. This overestimates the
effective sample size and underestimates the prediction error.

```{r random-cv-inflation}
# Simulate spatial data and diagnose its dependency structure
sim <- borg_simulate(n = 200, type = "spatial", autocorrelation = 0.7, seed = 42)

diagnosis <- borg_diagnose(sim$data, coords = c("x", "y"), target = "y.1",
                           verbose = FALSE)
cat("Dependency:", diagnosis@dependency_type, "\n")
cat("Recommended CV:", diagnosis@recommended_cv, "\n")
```

**Fix**: Use `borg_cv()` to generate appropriate blocked CV folds that
respect the dependency structure, or call `borg()` for a full automated
pipeline.


# Risk Interactions

In practice, evaluation risks rarely appear in isolation. Real-world datasets
often exhibit multiple dependency structures simultaneously, and their effects
compound in ways that are worse than the sum of parts.

## Spatial + Target Leakage

When a dataset has both spatial autocorrelation and a leaked feature, the two
risks interact. Spatial proximity allows the model to exploit local patterns,
while the leaked feature provides a near-perfect shortcut to the target. Random
CV inflates metrics through both channels at once. The combined inflation is
often much larger than either source alone.

```{r combined-risks}
# Simulate data with both spatial autocorrelation and target leakage
sim <- borg_simulate(n = 200, type = "combined", autocorrelation = 0.7,
                     leak_strength = 0.95, seed = 42)

# Random split
set.seed(1)
train_idx <- sample(200, 140)
test_idx <- setdiff(1:200, train_idx)

# Inspect for both risks
result <- borg_inspect(sim$data, train_idx = train_idx, test_idx = test_idx,
                       target = "y.1", coords = c("x", "y"))
result
```

In this example, `borg_inspect()` detects both the target leakage and the
spatial proximity issue. Fixing only one of them still leaves the evaluation
compromised.

## Preprocessing + Spatial Structure

Preprocessing leakage (fitting normalization on the full dataset) interacts
with spatial structure in a subtle way. When the train and test regions have
different covariate distributions (which is common in spatial data, where
western and eastern regions may differ in elevation, rainfall, or land cover),
the test region's statistics contaminate the training normalization. This makes
the model's predictions in the test region artificially better, because the
preprocessing has already "seen" the test distribution.

## Groups + Temporal Ordering

In longitudinal studies, group leakage and temporal violations can co-occur.
If patient A has observations at times t=1,2,3 and you randomly split, some
of patient A's early observations may end up in the test set while later
observations are in training. The model then predicts the past for a patient
it has already seen in the future. Both risks need to be addressed: split by
groups first, then ensure temporal ordering within each group.


# Decision Tree: Which Risks to Check

Not every dataset requires checking all risk types. Use this decision tree to
identify which risks are relevant for your data.

**Step 1: Data structure**

- Does your data have repeated measurements from the same subject, site, or
  unit? Check **group leakage**.
- Does your data have coordinates (longitude/latitude, UTM, or any x/y)?
  Check **spatial proximity** and **spatial overlap**.
- Does your data have a time column (dates, timestamps, or ordered sequence)?
  Check **temporal ordering**.

**Step 2: Preprocessing pipeline**

- Did you normalize, impute, or reduce dimensionality on the full dataset
  before splitting? Check **preprocessing leakage**.
- Did you fit a PCA, encoder, or embedding on all data? Same risk.

**Step 3: Feature provenance**

- Were any features computed from the target variable? Check **target
  leakage**.
- Were any features aggregated across time periods that include the test
  period? Check **target leakage** (temporal variant).
- Do any features have suspiciously high correlation with the target?
  Use `borg_inspect()` with the `target` argument to scan automatically.

**Step 4: Split mechanics**

- Did you create train/test splits manually? Check **index overlap**.
- Did your data go through deduplication after splitting? Check **duplicate
  rows**.
- Are you using nested CV? Check **CV fold contamination**.
- Did you fit the model separately from splitting? Check **model scope**.

**Quick start**: if you are unsure, run `borg_diagnose()` on your data to
detect dependency structure, then `borg_inspect()` on your split to scan for
leakage:

```{r decision-tree-example}
# Step 1: Diagnose the dataset
sim <- borg_simulate(n = 200, type = "spatial", autocorrelation = 0.5, seed = 1)

diagnosis <- borg_diagnose(sim$data, coords = c("x", "y"), target = "y.1",
                           verbose = FALSE)
cat("Structure:", diagnosis@dependency_type, "\n")
cat("Severity:", diagnosis@severity, "\n")

# Step 2: Inspect a candidate split
set.seed(1)
train_idx <- sample(200, 140)
test_idx <- setdiff(1:200, train_idx)

inspection <- borg_inspect(sim$data, train_idx = train_idx, test_idx = test_idx,
                           coords = c("x", "y"), target = "y.1")
cat("\nValid split?", inspection@is_valid, "\n")
cat("Hard violations:", inspection@n_hard, "\n")
cat("Soft warnings:", inspection@n_soft, "\n")
```

When `borg_diagnose()` reports spatial or temporal structure and
`borg_inspect()` flags spatial proximity on a random split, the diagnosis and
inspection agree: you need blocked CV.


# Quick Reference

| Risk Type | Severity | Detection Method | Fix |
|-----------|----------|------------------|-----|
| `index_overlap` | Hard | Index intersection | Use `setdiff()` |
| `duplicate_rows` | Hard | Row hashing | Deduplicate or group |
| `preprocessing_leak` | Hard | Parameter comparison | Fit on train only |
| `target_leakage` | Hard | Correlation > 0.99 | Remove feature |
| `group_leakage` | Hard | Group intersection | Group-aware split |
| `temporal_leak` | Hard | Timestamp comparison | Chronological split |
| `cv_contamination` | Hard | Fold index check | Rebuild folds |
| `model_scope` | Hard | Row count | Refit on train only |
| `proxy_leakage` | Soft | Correlation 0.95-0.99 | Domain review |
| `spatial_proximity` | Soft | Distance check | Spatial blocking |
| `spatial_overlap` | Soft | Convex hull | Geographic split |


# Accessing Risk Details

```{r risk-access}
# Create result with violations
result <- borg_inspect(
  data.frame(x = 1:100, y = rnorm(100)),
  train_idx = 1:60,
  test_idx = 51:100
)

# Summary
cat("Valid:", result@is_valid, "\n")
cat("Hard violations:", result@n_hard, "\n")
cat("Soft warnings:", result@n_soft, "\n")

# Individual risks
for (risk in result@risks) {
  cat("\n", risk$type, "(", risk$severity, "):\n", sep = "")
  cat("  ", risk$description, "\n")
  if (!is.null(risk$affected_indices)) {
    cat("  Affected:", head(risk$affected_indices, 5), "...\n")
  }
}

# Tabular format
as.data.frame(result)
```

## See Also

- `vignette("quickstart")` - Basic usage

- `vignette("frameworks")` - Framework integration
