---
title: "Temporal Cross-Validation"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
library(BORG)
library(ggplot2)

theme_borg <- theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)
  )
theme_set(theme_borg)
```

# Introduction

Time series cross-validation is one of those techniques that most practitioners
know they should use but often skip in favour of random k-fold CV. The cost of
that shortcut is inflated performance estimates. A model that appears to explain
60% of variance under random CV might explain 25% when evaluated honestly on
future data. The gap between those numbers is real: it shows up in production,
in failed forecasts, and in published results that do not replicate.

The problem is look-ahead bias. When random CV scatters observations into folds
without regard for time, the training set routinely contains future observations.
A model trained on data from January through December and tested on March has
already seen April. The test error shrinks because the model has access to
information that would not exist at prediction time. This is not a subtle
statistical nuance; it is a data leak that invalidates the entire evaluation.

The mechanism is straightforward. Suppose we want to predict species abundance
in July 2023. Under random 5-fold CV, the training set could include
observations from June and August 2023. A model that simply memorizes local
trajectories will report low test error because it interpolates between
temporally adjacent training points. Under temporal CV, the training set
contains only data from before July 2023. The model must extrapolate from
the past, which is the actual task we care about.

Look-ahead bias affects any dataset where observations have a natural temporal
ordering: financial time series, ecological monitoring records, climate
projections, sensor readings, patient health trajectories. In finance, a
stock-price model evaluated with random CV can appear to predict next-day
returns with 2% accuracy when the true out-of-sample accuracy is 0.5%. In
ecology, species distribution models fitted to multi-year survey data
routinely show R-squared inflation of 15-40% under random CV because annual
population counts are autocorrelated. In clinical medicine, patient health
trajectories measured at weekly intervals carry enough autocorrelation that
random CV inflates the apparent predictive power of longitudinal biomarker
models by 10-30%.

Panel data deserves special mention. Longitudinal studies track the same
units (patients, field sites, countries) over time. Both temporal
autocorrelation within each unit and similarity between repeated measurements
on the same unit inflate metrics simultaneously. A patient's blood glucose at
time $t$ predicts their blood glucose at time $t+1$, and a random CV fold
that places both observations in different sets allows the model to exploit
this identity-level autocorrelation. Ignoring either the temporal or the
group structure produces optimistic estimates; ignoring both produces
estimates that can be off by a factor of two.

BORG provides temporal cross-validation strategies that enforce chronological
ordering, insert embargo gaps between training and test periods, and detect
when random CV would produce misleading results. This vignette walks through
the full workflow: diagnosing temporal structure, choosing between blocking
and expanding window strategies, setting embargo periods, comparing random vs
temporal CV, and detecting concept drift in deployment data.

The functions we will use are `borg_simulate()`, `borg_diagnose()`, `borg_cv()`,
`borg_temporal_cv()`, `borg_compare_cv()`, `borg_drift()`, and
`borg_fold_performance()`. All code runs end to end with no external
dependencies beyond BORG and ggplot2.

# The problem: temporal autocorrelation

Temporal autocorrelation is the tendency for consecutive observations to
resemble each other. Today's temperature predicts tomorrow's. This quarter's
GDP predicts next quarter's. A patient's blood pressure reading at 10:00 AM
carries information about the reading at 10:15 AM. The closer two observations
are in time, the more correlated they tend to be.

Three quantities characterize the severity of temporal autocorrelation for
cross-validation purposes.

**Autocorrelation function (ACF).** The ACF measures the correlation between a
time series and lagged versions of itself. The lag-1 ACF is the correlation
between consecutive observations. For financial returns, this might be 0.05
(nearly independent). For daily temperature, it exceeds 0.9. For ecological
population counts measured annually, values of 0.4-0.7 are common. When the
lag-1 ACF exceeds 0.2, random CV starts to produce measurably inflated
estimates.

**Ljung-Box test.** The Ljung-Box Q-statistic tests the null hypothesis that
the first *k* autocorrelations are all zero. A significant result (p < 0.05)
means the series contains temporal structure that CV must respect. BORG
computes this at lag 10 by default. A non-significant result does not
guarantee independence (the series could have nonlinear dependence), but a
significant result is a clear signal that random CV is inappropriate.

**Effective sample size.** With autocorrelation, 500 observations do not carry
as much information as 500 independent observations. The effective sample size
is $n_{\text{eff}} = n \cdot (1 - \rho) / (1 + \rho)$, where $\rho$ is the
lag-1 ACF. A series of 500 observations with $\rho = 0.8$ has an effective
sample size of about 56. Random CV acts as if each of those 500 observations
contributes independent test information. They do not. The test error is
estimated from 56 effective data points masquerading as 500, producing
confidence intervals that are far too narrow and point estimates that are
biased toward optimism.

The connection between autocorrelation and CV inflation is mechanical.
Consider a simple scenario: you want to predict $y_{t+1}$ given predictors
$x_t$. In random CV, the training set might contain $y_{t-1}$, $y_t$,
$y_{t+2}$, and $y_{t+3}$. If $\rho = 0.8$, the model can predict $y_{t+1}$
to within $\sqrt{1 - \rho^2} \approx 0.6$ standard deviations just by
interpolating between the surrounding training points. The apparent R-squared
rises, but the model has learned nothing about the underlying data-generating
process; it has memorized the local trajectory.

The interaction between autocorrelation and model complexity matters. Linear
models are somewhat protected because they cannot memorize local trajectories
as aggressively; they fit a global hyperplane and the autocorrelation mainly
inflates confidence intervals rather than point predictions. Tree-based models
(random forests, gradient boosted trees) are more vulnerable because they
partition the feature space into small regions and can effectively memorize
clusters of temporally adjacent training points. A random forest trained with
random CV on daily temperature data will place nearby days in the same leaf
nodes and report near-perfect test accuracy. The same forest evaluated with
temporal CV will show the true predictive skill, which for a naive model is
often poor. Neural networks sit somewhere in between: their capacity to
memorize depends on architecture and regularization, but high-capacity
networks with dropout disabled can exploit temporal leakage as aggressively
as tree-based methods.

The character of the autocorrelation also matters. Strong short-range
autocorrelation, like daily weather or hourly sensor data, means that
observations 1-3 time steps apart are nearly identical. The lag-1 ACF might
be 0.95. Random CV on such data is almost equivalent to testing on the
training set itself, because adjacent observations end up on opposite sides
of the train/test split. The inflation is large (often 30-60%) but easy to
detect with `borg_diagnose()`. Weak long-range autocorrelation is subtler.
Annual ecological surveys with lag-1 ACF of 0.3-0.5 produce moderate
inflation (5-20%) that is easy to dismiss as noise. But at these
autocorrelation levels, the effective sample size is already 40-70% of the
nominal $n$, and confidence intervals around the CV estimate are meaningfully
too narrow. The danger with weak long-range autocorrelation is not that the
point estimate is wildly wrong, but that the practitioner believes the
estimate is more precise than it is.

# Simulating temporal data

We start by generating data with known temporal structure. `borg_simulate()`
with `type = "temporal"` creates AR(1) predictors and an AR(1) noise process.
Both the features and the response exhibit temporal autocorrelation, which is
the typical scenario in applied work where predictors are themselves time
series (economic indicators, environmental measurements, lagged variables).

```{r simulate}
sim <- borg_simulate(
  n = 500,
  type = "temporal",
  n_predictors = 5,
  signal_strength = 0.3,
  autocorrelation = 0.7,
  seed = 42
)

sim
```

The simulated data contains a `time` column (daily dates starting from
2020-01-01), five predictors (`x1` through `x5`), and a response `y`. The
true R-squared is `r round(sim$true_r2, 2)`, which represents the ceiling
for any model that correctly recovers the signal without temporal leakage.

```{r colnames}
names(sim$data)
```

We can visualize the response over time to see the autocorrelation structure:

```{r plot-ts, fig.width = 7, fig.height = 4}
ggplot(sim$data, aes(x = time, y = y)) +
  geom_line(alpha = 0.6) +
  labs(title = "Simulated temporal response (AR(1), rho = 0.7)",
       x = "Time", y = "Response")
```

The smooth, slowly varying trajectory is the visual fingerprint of temporal
autocorrelation. Consecutive values track each other closely, which means
that a random train/test split would place temporally adjacent observations
on both sides of the partition.

Now we confirm the temporal structure formally with `borg_diagnose()`:

```{r diagnose}
diag <- borg_diagnose(
  sim$data,
  time = "time",
  target = "y",
  verbose = FALSE
)

diag
```

BORG detects the temporal dependency, reports the lag-1 ACF, the Ljung-Box
p-value, the decorrelation lag, and the minimum embargo period. The
`@recommended_cv` slot tells us which CV strategy BORG would select
automatically.

The diagnostic output deserves careful reading. The lag-1 ACF is the
single most informative number: values above 0.2 indicate that random CV
will produce measurably inflated estimates, and values above 0.5 indicate
that the inflation will be large enough to change substantive conclusions.
The Ljung-Box p-value is a formal hypothesis test; a value below 0.05
rejects the null of independence, but values well below 0.001 (as we see
here) indicate that the temporal structure is not borderline but
unambiguous. The decorrelation lag tells us how many time steps it takes
for the autocorrelation to decay below the significance threshold
($2 / \sqrt{n}$). For our data with $\rho = 0.7$ and $n = 500$, the
significance threshold is about 0.09, and the ACF at lag $k$ is
approximately $0.7^k$, so it takes roughly 27 time steps for the
autocorrelation to become negligible. The embargo minimum is derived from
this decorrelation lag and represents the smallest gap between training
and test sets that will prevent temporal leakage. Setting the embargo
below this value leaves residual autocorrelation between the last
training observation and the first test observation.

We can inspect the temporal diagnostics directly:

```{r temporal-slots}
cat("ACF lag-1:       ", diag@temporal$acf_lag1, "\n")
cat("Ljung-Box p:     ", diag@temporal$ljung_box_p, "\n")
cat("Embargo minimum: ", diag@temporal$embargo_minimum, "\n")
```

The lag-1 ACF of `r round(diag@temporal$acf_lag1, 3)` confirms strong temporal
dependence. The Ljung-Box p-value is effectively zero, rejecting the null of
no autocorrelation. The embargo minimum tells us how many time units we need
between training and test sets to break the autocorrelation link.

# Temporal block CV

Temporal block CV divides the time series into contiguous blocks and assigns
each block to a fold. Within each iteration, the test set is a continuous time
window, and the training set is everything before it (minus an embargo gap).
This mimics the real prediction task: we train on the past and predict the
future.

```{r temporal-block}
cv_block <- borg_cv(
  sim$data,
  time = "time",
  target = "y",
  strategy = "temporal_block",
  v = 5,
  verbose = TRUE
)

cv_block
```

The fold structure preserves chronological order. Each fold tests on a later
time window than the previous fold's test set. Training data is always
temporally prior to test data, with an embargo gap in between.

We can verify the chronological structure by checking the time ranges of
training and test sets in each fold:

```{r fold-ranges}
for (i in seq_along(cv_block$folds)) {
  f <- cv_block$folds[[i]]
  train_range <- range(sim$data$time[f$train])
  test_range <- range(sim$data$time[f$test])
  cat(sprintf("Fold %d: train %s to %s | test %s to %s\n",
              i, train_range[1], train_range[2],
              test_range[1], test_range[2]))
}
```

In each fold, the training period ends before the test period begins. The gap
between the last training observation and the first test observation is the
embargo period. No future data leaks into the training set.

The contiguous block structure is not just a convenience; it is a requirement.
Consider the alternative: assigning random time points to folds while
keeping folds "mostly" in order. A fold that tests on March 2021 but
includes two February 2021 observations in its test set and three March 2021
observations in the training set has a subtle leak. The autocorrelation
between the March training and March test observations acts as a bridge.
Contiguous blocks eliminate this by construction: every observation in the
test fold is later than every observation in the training set. There is no
ambiguity and no possibility of partial leakage.

We can also use the convenience wrapper `borg_temporal_cv()`, which returns
an rsample-compatible `borg_rset` object for use with tidymodels:

```{r temporal-cv-wrapper, eval = requireNamespace("rsample", quietly = TRUE)}
cv_rset <- borg_temporal_cv(
  sim$data,
  time = "time",
  target = "y",
  v = 5
)

cv_rset
```

This plugs directly into `tune::tune_grid()` and `tune::fit_resamples()`
without any adapter code.

Per-fold performance gives us a direct view of how model accuracy changes
over time:

```{r fold-perf}
perf <- borg_fold_performance(
  sim$data,
  folds = cv_block,
  formula = y ~ x1 + x2 + x3 + x4 + x5,
  metric = "rmse"
)

perf
```

The per-fold RMSE values carry diagnostic information that aggregate metrics
hide. Three patterns are worth looking for. First, a monotonically increasing
RMSE across folds suggests concept drift: the relationship between predictors
and response is changing over time, and the model trained on early data
predicts later periods poorly. Second, a U-shaped pattern (high RMSE in early
and late folds, low in the middle) can indicate seasonality or a periodic
process that the model captures during its central period but misses at the
edges. Third, a single spike in one fold often points to an anomalous event
(a sensor failure, a policy change, an extreme weather year) rather than a
systematic problem. Random CV averages all of these patterns away, producing
a single number that hides the temporal structure of model performance.

# Expanding window CV

Temporal block CV assigns each time block to exactly one fold. Expanding
window CV (also called forward chaining or walk-forward validation) takes a
different approach: it grows the training set over time. Fold 1 trains on the
first block and tests on the second. Fold 2 trains on the first two blocks and
tests on the third. By the final fold, the training set contains almost all
historical data.

```{r expanding}
cv_expand <- borg_cv(
  sim$data,
  time = "time",
  target = "y",
  strategy = "temporal_expanding",
  v = 5,
  verbose = TRUE
)

cv_expand
```

Compare the training set sizes across folds:

```{r expanding-sizes}
sizes <- data.frame(
  fold = seq_along(cv_expand$folds),
  n_train = vapply(cv_expand$folds, function(f) length(f$train), integer(1)),
  n_test = vapply(cv_expand$folds, function(f) length(f$test), integer(1))
)

sizes
```

The training set grows with each fold while the test set remains roughly
constant. This reflects the real deployment scenario more faithfully than block
CV when the model will be retrained on all available data before making
predictions.

**When to use which.** Temporal block CV is appropriate when you want
unbiased estimates of average future performance and when the data-generating
process is approximately stationary. Each fold contributes roughly equal
weight, and the estimate is not dominated by the largest training set.
Expanding window CV is appropriate when the model will be deployed with a
growing training history (which is almost always the case in production
forecasting). It answers the question: "Given everything I know up to time
$t$, how well do I predict the next window?" The cost is that early folds have
small training sets and may underperform, which inflates the variance of the
overall estimate.

The bias-variance tradeoff between the two strategies is concrete. Temporal
block CV uses a fixed training set size for each fold (roughly $n \times
(v-1)/v$ observations). This gives each fold comparable statistical power, so
the per-fold estimates have similar variance. But if the data-generating
process evolves over time, the early folds train on data that may no longer
represent the current regime, introducing bias. Expanding window CV reduces
bias in later folds because it trains on the most data, but early folds
suffer from high variance due to small training sets. In practice, the
expanding window estimate is dominated by the last few folds (which have
the most training data and the most test-relevant information), while the
block CV estimate weights all folds equally.

A third option exists: sliding window CV (also called rolling window). Here,
the training set has a fixed width and slides forward in time. Fold 1 trains
on observations 1-100 and tests on 101-120. Fold 2 trains on 21-120 and
tests on 121-140. The training set never grows; old data is dropped as new
data enters. This is the right strategy when old data is actively harmful
(e.g., a financial regime changed two years ago, and pre-change data misleads
the model). BORG does not currently implement sliding window CV as a separate
strategy, but you can approximate it by setting `strategy = "temporal_block"`
and filtering the training indices in each fold to include only observations
within a fixed time window before the test period.

A practical heuristic: if you have fewer than 200 observations, use expanding
window CV to avoid wasting data on small training sets. If you have 500+
observations and care about average performance across the full time range,
use temporal block CV. If concept drift is a concern, expanding window CV
is more informative because you can plot performance against time and see
whether the model degrades.

# Embargo periods

The embargo parameter controls the gap between the last training observation
and the first test observation. Without an embargo, the observation immediately
before the test window appears in training. If autocorrelation is high, that
observation carries information about the test data, and the information leak
(while reduced compared to random CV) persists.

BORG computes a recommended embargo from the decorrelation lag: the number of
time steps after which the ACF drops below the significance threshold
($2 / \sqrt{n}$). For our simulated data with $\rho = 0.7$, the decorrelation
lag is several time steps, and the embargo needs to span at least that many
time steps to eliminate residual leakage.

Here we compare temporal block CV with and without an embargo:

```{r embargo-compare}
cv_no_embargo <- borg_cv(
  sim$data,
  time = "time",
  target = "y",
  strategy = "temporal_block",
  embargo = 0,
  v = 5,
  verbose = FALSE
)

cv_embargo <- borg_cv(
  sim$data,
  time = "time",
  target = "y",
  strategy = "temporal_block",
  embargo = 10,
  v = 5,
  verbose = FALSE
)
```

Check training set sizes to see the effect of the embargo:

```{r embargo-sizes}
no_emb_sizes <- vapply(cv_no_embargo$folds,
                       function(f) length(f$train), integer(1))
emb_sizes <- vapply(cv_embargo$folds,
                    function(f) length(f$train), integer(1))

cat("Without embargo:", paste(no_emb_sizes, collapse = ", "), "\n")
cat("With embargo=10:", paste(emb_sizes, collapse = ", "), "\n")
```

The embargo removes training observations that fall within the gap window.
Larger embargoes exclude more data, which reduces the effective training set
but also reduces leakage. The trade-off is clear: too small and leakage
persists; too large and training data is wasted.

The embargo must also account for feature engineering. If the model uses
lagged features or rolling statistics, the embargo needs to be at least as
wide as the widest feature window. Consider a model that uses 7-day rolling
averages as input features. The rolling average at day $t$ is computed from
days $t-6$ through $t$. If the embargo is only 3 days, the training set
includes observations from day $t-3$, and the 7-day rolling average at day
$t$ in the test set was computed using data from day $t-6$, which overlaps
with the training period. The embargo must be at least 7 days to prevent
this indirect leakage. More generally, if the model uses $k$-step lagged
features, rolling windows of width $w$, or exponentially weighted moving
averages with a half-life of $h$, the embargo should be at least
$\max(k, w, 3h)$ time steps. BORG's automatic embargo is based on the
raw autocorrelation of the response, not on the feature engineering
pipeline, so you may need to override it with a larger value when lagged
or rolling features are present.

How to choose the right embargo:

- Set it to the decorrelation lag from `borg_diagnose()`. This is the default
  when you let BORG choose automatically.
- For daily financial data, 1-5 days is typical.
- For monthly ecological monitoring, 2-6 months covers most autocorrelation
  structures.
- For annual data with low sample sizes, an embargo of 1-2 years may consume
  too many observations. In that case, accept a slightly optimistic estimate
  rather than discarding 20% of your data as an embargo buffer.

If you are uncertain, run `borg_fold_performance()` with several embargo values
and check whether the per-fold RMSE stabilizes. If increasing the embargo from
5 to 10 does not change the metrics, 5 is sufficient.

# Comparing random vs temporal CV

The strongest evidence for using temporal CV is an empirical comparison on
your own data. `borg_compare_cv()` runs both random and blocked CV, repeats
the comparison for stability, and tests whether the performance gap is
statistically significant.

```{r compare, fig.width = 7, fig.height = 5}
comparison <- borg_compare_cv(
  sim$data,
  formula = y ~ x1 + x2 + x3 + x4 + x5,
  time = "time",
  metric = "rmse",
  v = 5,
  repeats = 10,
  seed = 42,
  verbose = FALSE
)

comparison
```

The output reports mean RMSE under random and temporal CV, the estimated
inflation as a percentage, and a p-value from a paired t-test. For our
simulated data with $\rho = 0.7$, random CV typically underestimates RMSE
by 20-40%. That gap represents the look-ahead bias.

Visualize the comparison:

```{r plot-compare, fig.width = 6, fig.height = 5}
plot(comparison, type = "boxplot")
```

Each box shows the distribution of RMSE values across repeats. When the boxes
do not overlap, the inflation is both statistically and practically
significant. This plot serves as direct evidence for reviewers, or for
yourself, that random CV on this dataset gives a misleading picture of model
performance.

To calibrate our intuition, we can run the same comparison on independent
data where random CV should be fine:

```{r compare-independent}
sim_ind <- borg_simulate(
  n = 500,
  type = "independent",
  n_predictors = 5,
  signal_strength = 0.3,
  seed = 123
)

comp_ind <- borg_compare_cv(
  sim_ind$data,
  formula = y ~ x1 + x2 + x3 + x4 + x5,
  metric = "rmse",
  v = 5,
  repeats = 10,
  seed = 123,
  verbose = FALSE
)

comp_ind
```

With independent data, the two CV approaches should give similar RMSE values.
Any residual gap is due to the different fold geometries (temporal folds are
contiguous; random folds are scattered), not leakage. This sanity check
confirms that temporal CV is not introducing unnecessary pessimism when the
data are genuinely independent.

The inflation number from `borg_compare_cv()` belongs in your methods section.
A statement like "Random 5-fold CV underestimated RMSE by 32% relative to
temporal block CV (paired t-test, p < 0.001)" gives readers a concrete,
dataset-specific measure of how much the evaluation strategy matters. For a
journal publication, we recommend including the following numbers: the mean
and standard deviation of the metric under both random and temporal CV, the
inflation percentage, and the paired t-test p-value. Reviewers increasingly
expect this comparison, especially in ecology and environmental science where
Roberts et al. (2017) has made temporal and spatial CV standard practice.

How you respond to the inflation result depends on its magnitude. If the
inflation is small (below 5%), the temporal structure in your data is weak
enough that random CV gives a reasonable estimate. You can note this in the
methods section and proceed with either strategy. If the inflation is between
5% and 20%, random CV is measurably optimistic but the qualitative
conclusions (which model is best, which features matter) are likely
unchanged. Report the temporal CV metrics as the primary result and mention
the inflation as context. If the inflation exceeds 20%, random CV is
actively misleading. The model's apparent performance is substantially
better than its true out-of-sample performance. In this regime, random CV
should not be used for model selection, hyperparameter tuning, or
performance reporting. Use temporal CV exclusively, and flag the inflation
to co-authors and reviewers as evidence that the evaluation strategy is
not a methodological detail but a first-order concern.

# Concept drift detection

Temporal CV produces honest estimates of how well a model predicts the near
future. But what happens when the deployment data differs from the training
data? In time series, the data-generating process can change: a climate
regime shifts, a market enters a new phase, a population crosses a tipping
point. This is concept drift, and it means that even honestly estimated
performance on historical data overestimates performance on deployment data.

`borg_drift()` detects whether the distribution of predictor variables has
shifted between a training set and a new dataset. It combines per-feature
Kolmogorov-Smirnov tests with a multivariate classifier two-sample test.

To demonstrate, we split our simulated data into an early training period
and a late deployment period:

```{r drift-split}
n <- nrow(sim$data)
train_data <- sim$data[1:round(n * 0.7), ]
deploy_data <- sim$data[(round(n * 0.7) + 1):n, ]

cat("Training period:",
    as.character(min(train_data$time)), "to",
    as.character(max(train_data$time)), "\n")
cat("Deployment period:",
    as.character(min(deploy_data$time)), "to",
    as.character(max(deploy_data$time)), "\n")
```

Now test for drift:

```{r drift-test}
drift <- borg_drift(
  train = train_data,
  new = deploy_data,
  predictors = paste0("x", 1:5),
  seed = 42
)

drift
```

For our AR(1) simulation with a constant data-generating process, the drift
test should report no severe shift: the predictor distributions are
stationary by construction. The classifier AUC should be near 0.5 (no better
than random at distinguishing training from deployment), and the per-feature
KS tests should mostly be non-significant.

In real applications, concept drift arises from identifiable causes. Regime
changes are the most common: a lake transitions from oligotrophic to
eutrophic, an economy enters a recession, or a new policy alters the
relationship between predictors and outcomes. Sensor degradation is another
source: an instrument that drifts out of calibration over months produces
features whose distributions shift gradually. Seasonal or cyclical processes
can mimic drift if the training period does not cover a full cycle. A model
trained on spring and summer data will see apparent drift when evaluated on
autumn data, even though the underlying process is stationary over years.
Finally, changes in data collection protocols (a new survey method, a
different sampling frequency, a replaced sensor) create abrupt distributional
shifts that are easy to detect but hard to correct without retraining.

Here is an artificial example where we shift the deployment data:

```{r drift-shifted}
deploy_shifted <- deploy_data
deploy_shifted$x1 <- deploy_shifted$x1 + 1.5
deploy_shifted$x2 <- deploy_shifted$x2 * 2

drift_shifted <- borg_drift(
  train = train_data,
  new = deploy_shifted,
  predictors = paste0("x", 1:5),
  seed = 42
)

drift_shifted
```

With two features shifted, the classifier AUC rises well above 0.5, and the
per-feature results identify exactly which variables drifted. The
`overall_severity` field summarizes whether the shift is mild, moderate, or
severe.

The severity levels map to concrete actions. "None" means the deployment data
is indistinguishable from the training data; no action is needed. "Mild"
means one or two features show statistically significant shifts, but the
effect sizes are small (KS statistic below 0.15). The model's predictions
are likely still reliable, but monitoring the drifted features over the next
evaluation cycle is prudent. "Moderate" means the classifier can distinguish
training from deployment data with AUC above 0.65, or multiple features show
KS statistics between 0.15 and 0.30. Performance degradation is likely, and
retraining on more recent data is recommended before deployment. "Severe"
means the classifier AUC exceeds 0.75 or at least one feature has a KS
statistic above 0.30. The deployment data comes from a different regime than
the training data, and any performance estimates derived from the training
period are unreliable. Retrain the model, investigate which features changed
and why, and consider whether the model's feature set is still appropriate
for the new regime.

The practical workflow is:

1. Train your model and evaluate it with temporal CV.
2. Before deploying, run `borg_drift()` on the deployment data.
3. If severity is "none" or "mild", proceed. If "moderate" or "severe",
   retrain on more recent data or investigate which features changed and why.

Concept drift detection does not replace temporal CV; it extends it. Temporal
CV tells you how well the model predicts the next time window given the
current training data. Drift detection tells you whether the next time window
is similar enough to the training data for those predictions to be trustworthy.

# Practical guidance

## Choosing a strategy

The choice between temporal block CV and expanding window CV depends on the
prediction task and dataset size.

**Temporal block CV** (`strategy = "temporal_block"`): the default for
temporal data. Each fold tests on a contiguous time block. Training data
comes from before the test block. Use this when you want an average estimate
of future prediction performance and when the time series is long enough
(300+ observations with 5 folds, so each fold has 60+ observations).

**Expanding window CV** (`strategy = "temporal_expanding"`): the training
set grows with each fold. Use this when the model will be retrained on all
available data before deployment. Also the better choice for shorter time
series (100-300 observations) because early folds still have enough training
data to fit a reasonable model.

**Purged CV** (`strategy = "purged_cv"`): de Prado's combinatorial purged
cross-validation. Uses all combinations of time groups as test sets, with
embargo gaps. Most conservative; computationally expensive. Use for financial
applications where even small leakage has large dollar consequences.

A decision tree:

- Fewer than 100 observations -> expanding window with v = 3.
- 100-300 observations -> expanding window with v = 5.
- 300+ observations, stationary process -> temporal block with v = 5.
- 300+ observations, suspected drift -> expanding window with v = 5-10,
  then plot per-fold performance to detect degradation.
- Financial time series -> purged CV with appropriate embargo.

## Rules of thumb with specific numbers

**Minimum dataset size.** Temporal block CV with 5 folds requires at least
100 observations for each fold to have 20 test observations. Below 50 total
observations, expanding window with v = 3 is the practical minimum.

**Embargo size.** Set it to the decorrelation lag. For daily data with
$\rho = 0.7$, this is typically 5-10 days. For monthly data with $\rho = 0.5$,
2-4 months. The embargo should never exceed 20% of the total time range;
if it does, the training sets become too small.

**Number of folds.** Five is the default. Three folds work for short series
(under 200 observations). Ten folds are appropriate for long series (1000+
observations) where you want lower-bias estimates.

**Repeats.** Temporal block CV is deterministic given the data ordering, so
repeats = 1 is fine for a single run. Use repeats = 5-10 in `borg_compare_cv()`
because the random CV baseline needs repeated sampling for stability.

**When to worry about inflation.** If `borg_compare_cv()` shows less than
5% inflation, temporal CV and random CV are practically equivalent for your
data. This can happen when the autocorrelation is weak ($\rho < 0.1$) or
when the signal-to-noise ratio is high enough that the signal dominates
the autocorrelation structure.

**Feature engineering pipelines.** Temporal CV interacts with feature
engineering in ways that are easy to overlook. If features are computed
from the response variable (e.g., lagged response, rolling mean of
response), those computations must happen inside the CV loop, not before
it. Computing a 30-day rolling average of the response on the full dataset
and then splitting into temporal folds leaks future response values into
the training set through the rolling window. BORG does not perform feature
engineering, so this is the user's responsibility. The safe approach is to
compute time-dependent features separately for each fold's training set and
apply the same transformation (using only training-set parameters) to the
test set. For features derived only from predictors (e.g., lagged
temperature, rolling mean of precipitation), pre-computation on the full
dataset is safe as long as the embargo accounts for the feature window width.

**Reporting temporal CV in publications.** The methods section should state
the CV strategy (temporal block or expanding window), the number of folds,
the embargo size in the same time units as the data (e.g., "10-day embargo"
not "embargo = 10"), and whether the embargo was set automatically or
overridden. If `borg_compare_cv()` was run, report the inflation percentage
and the paired t-test p-value. Include per-fold metrics in a supplementary
table or figure so that reviewers can assess whether performance is stable
across time or shows evidence of drift. A sentence like "Temporal block CV
with 5 folds and a 10-day embargo produced mean RMSE of 0.42 (SD 0.05);
random 5-fold CV underestimated RMSE by 28% (paired t-test, p < 0.001)"
is sufficient for most journals.

## When NOT to use temporal CV

Temporal CV is not always the right choice, even when the data has timestamps.

**Interpolation tasks.** If the goal is to fill gaps in a time series (e.g.,
imputing missing sensor readings between known values), random CV correctly
mimics the prediction task. The model will interpolate at prediction time,
so testing its interpolation ability is appropriate.

**Cross-sectional data with timestamps.** If you survey 500 sites once, and
each observation has a sampling date, but the dates reflect logistics (which
field crew visited when) rather than temporal dynamics, the timestamps carry
no temporal autocorrelation. Use `borg_diagnose()` to check: if the Ljung-Box
test is non-significant and the lag-1 ACF is below 0.1, random CV is safe.

**Stationary data with weak autocorrelation.** When $\rho < 0.1$ and the
Ljung-Box test is non-significant, the series is effectively independent
for cross-validation purposes. Temporal blocking would reduce the effective
sample size without improving the estimate.

**Very short time series.** With fewer than 30 observations, no CV strategy
produces stable estimates. Consider a train/test split at a fixed time point,
or use bootstrap-based confidence intervals instead.

In all these cases, set `allow_random = TRUE` in `borg_cv()` to override the
automatic blocking. BORG will issue a warning with the estimated inflation, so
the decision is transparent and documented.

## Panel data

Panel data (multiple entities observed over time) requires both temporal
and group structure. Each entity (patient, site, country) has its own time
series, and observations within an entity are autocorrelated.

For panel data, pass both `time` and `groups` to `borg_cv()`:

```{r panel, eval = FALSE}
cv_panel <- borg_cv(
  panel_data,
  time = "date",
  groups = "site_id",
  target = "response",
  v = 5
)
```

BORG will detect the joint structure and select a strategy that respects both
the temporal ordering and the group boundaries. No site appears in both
training and test within a fold, and training observations are always
temporally prior to test observations.

If groups are not important (all entities share the same data-generating
process and you want to predict future values for the same entities),
temporal blocking alone is sufficient. If entities are important (you want
to predict for new entities), group CV is needed in addition to temporal
blocking.

## End-to-end workflow

Putting the pieces together, here is the workflow for temporal model
validation:

1. **Diagnose.** Run `borg_diagnose()` with the `time` argument. Check
   the lag-1 ACF, the Ljung-Box p-value, and the recommended embargo.
2. **Generate folds.** Call `borg_cv()` with `time = "time"`. Let BORG
   select the strategy automatically, or choose between `"temporal_block"`
   and `"temporal_expanding"` based on the decision tree above.
3. **Set the embargo.** The default (from diagnosis) is usually adequate.
   Override with `embargo = k` if you have domain knowledge about the
   decorrelation time.
4. **Evaluate per-fold.** Use `borg_fold_performance()` to compute
   per-fold metrics. Plot them in order. If performance degrades in later
   folds, concept drift may be present.
5. **Compare.** Run `borg_compare_cv()` at least once. Report the
   inflation percentage in your methods section.
6. **Check for drift.** Before deployment, run `borg_drift()` on the
   deployment data. Retrain if severity is moderate or above.
7. **Report.** Include the CV strategy, the embargo size, the number of
   folds, the inflation estimate, and the drift severity. These five
   numbers let readers assess whether the reported metrics are trustworthy.

```{r workflow-summary}
cat("Diagnosis:\n")
cat("  Dependency type:", diag@dependency_type, "\n")
cat("  Severity:       ", diag@severity, "\n")
cat("  Recommended CV: ", diag@recommended_cv, "\n\n")
cat("CV strategy used: ", cv_block$strategy, "\n")
cat("Number of folds:  ", length(cv_block$folds), "\n")
```

A few extra lines of code, a few extra numbers in the methods section. The
payoff is an evaluation you can trust, and a model that performs in deployment
the way your metrics predicted it would.

# References

- Roberts, D.R., et al. (2017). Cross-validation strategies for data with
  temporal, spatial, hierarchical, or phylogenetic structure. *Ecography*,
  40(8), 913-929. doi:10.1111/ecog.02881

- de Prado, M.L. (2018). *Advances in Financial Machine Learning*. John
  Wiley & Sons. Chapter 7: Cross-Validation in Finance.

- Bergmeir, C., & Benitez, J.M. (2012). On the use of cross-validation
  for time series predictor evaluation. *Information Sciences*, 191, 192-213.
  doi:10.1016/j.ins.2011.12.028

- Cerqueira, V., Torgo, L., & Mozetic, I. (2020). Evaluating time series
  forecasting models: An empirical study on performance estimation methods.
  *Machine Learning*, 109, 1997-2028. doi:10.1007/s10994-020-05910-7

- Tashman, L.J. (2000). Out-of-sample tests of forecasting accuracy: an
  analysis and review. *International Journal of Forecasting*, 16(4), 437-450.
  doi:10.1016/S0169-2070(00)00065-0
